---
title: "Examining PacBio data with pbbamr"
author: "Nigel Delaney, Yuan Tian"
date: '`r Sys.Date()`'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Examining Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This package examines aligned data in the PacBio BAM format.  It can be used to 
investigate the BAM pbi index file to get bulk alignment statistics, or can be
used for deeper analysis of individual alignments by directly accessing
the BAM file alignments.

## Package installation

```
install.packages("devtools")
library(devtools)
# Don't build the vignette by default, you can save time by viewing it online
install_github("PacificBiosciences/pbbamr", build_vignettes = FALSE)
```
Note that on some PacBio machines, a configuration conflict may lead to the
message `Problem with the SSL CA cert (path? access rights?)`.  If this appears,
simply run the following command before those given above:

    httr::set_config( httr::config( ssl_verifypeer = 0L ) )

## Loading and plotting pbi files

Simply pass a BAM (.bam) or Dataset (.xml) file name to the `loadPBI` function
which will return a data frame with the information available in the pbi file.

```{r, echo=TRUE, results='asis'}
library(pbbamr)
# For this example, we will load some sample data
# distributed with the package.
bamname = system.file("extdata", "bam_mapping_1.bam", package="pbbamr")
# Load the index table.
ind = loadPBI(bamname)
# Show the first 3 rows (transposed, note kable just formats)
knitr::kable(t(ind[1,]))
```

The fields in the data frame closely correspond to the [published documentation
of the
format](http://pacbiofileformats.readthedocs.org/en/latest/PacBioBamIndex.html).[^1]

[^1]: Note that because R does cannot represent long integers, we store the file offset
as a string (which is a detail you can ignore).


The PacBio index table can tell you a lot about the data.  For example we plot a
distribution of alignment length and accuracy below.

**Alignment Length Distributions**

```{r, echo=TRUE, results='asis'}
library(ggplot2)
# We can calculate the length of the alignment on the reference as 
# the template end minus the template start of that alignment.
ind$alnLength = ind$tend - ind$tstart
ggplot(ind, aes(x=alnLength)) + geom_density(fill="blue") + theme_bw() +
  labs(x="Alignment Length on Reference")
```

**Accuracy Distribution**

Accuracy is a metric with many possible definitions.  Below, we will examine an
error rate defined as the total number of errors per reference basepair.

```{r, echo=TRUE, results='asis'}
ind$errorRate = (ind$mismatches + ind$inserts + ind$dels) / ind$alnLength 
ggplot(ind, aes(x=errorRate)) + geom_density(fill="cyan") + theme_bw() +
  labs(x="Error Rate")
```


## Adding more covariates to the index table

The PBI data is very useful, however some valuable information is stored within 
the main records.  If desired, while loading the index you can pass in some 
flags to gain optional information directly from the BAM file.  The following
other pieces of information can be collected:

| Optionally Available Data  |
| ----  |
| Read Quality |
| Number of Passes |
| SNR-A |
| SNR-C |
| SNR-G |
| SNR-T |
| Scrap Annotation (Only available for scraps.bam files)|

And you can do this with simple flags:

```
ind <- loadPBI(bamname, loadSNR=TRUE, loadNumPasses = TRUE, loadRQ = TRUE)
# Note loading scraps files can only be done if the name ends with ".scraps.bam" as below
# ind <- loadPBI("myfile.scraps.bam", loadSC = TRUE)
```

If the data file does not have the requested fields, NA values will be provided 
instead.  Since parsing the BAM file for this data is more time consuming than
just reading the index, it is not loaded by default.


## Examining Individual Alignments

Given an index file, we may want to drill down and examine the individul
alignments.  `pbbamr` allows you to do this if you have an indexed fasta file of
the reference available.  The function `loadAlnsFromIndex` will take a list of
file offsets (from the pbi index file) and return each BAM record specified by a
file offset as a data frame with rows corresponding to alignment positions and
columns as read or reference bases.

```{r, echo=TRUE, results='hold'}
# Get the name of the sample indexed FASTA file
fastaname = system.file("extdata", "lambdaNEB.fa", package="pbbamr")
# let's just grab and plot one alignment.
allAlns = loadAlnsFromIndex(ind, fastaname, 1)
# The alignments are returned as a list of data frames
# in this case let's just look at the first
aln = allAlns[[1]]
head(aln[1:6,])
```

Each alignment (or BAM record) is returned as a data frame element of a list. 
With this, one could combine all of these into one super alignment to gain
statistics.  For example, let's validate that the number of insertions in the
alignments matches the pbi file.


```{r, echo=TRUE, collapse=TRUE, results='hold'}
# Get all the alignments for all records in the index
allAlns = loadAlnsFromIndex(ind, fastaname)
# let's combine the individual alignments into one big data frame
alns = do.call(rbind, allAlns)
# Now count the number of insertions
insert_cnt = sum(alns$ref == "-")
# Do they match the pbi file?
insert_cnt == sum(ind$inserts)
```

More usefully though, one may want to perform some operation over a set of 
alignments to gather some summary statistics from them.  For example, let's 
write a cheesy function to figure out for single deletion events which basepair
was deleted and which basepair followed that deletion.

```{r, echo=TRUE, results='asis'}
# First let's write a function to find single deletion events, and report 
# the deleted base and following base.
findSingleDeletions <- function(aln) {
  # Where are the deletions of size 1?
  # First we will convert it to run length encoding (gap = 5)
  rls = rle(as.integer(aln$read))
  # Now find out what has a deletion of size 1
  del_size1 = which(rls$values == 5 & rls$lengths == 1)
  # and convert back from run length to standard coordinates
  org_positions = cumsum(rls$lengths)
  single_del_locs = org_positions[del_size1]
  # Presumably the alignment should not end with 
  # a deletion, but let's be sure
  single_del_locs = single_del_locs[single_del_locs < nrow(aln)]
  # And now let's grab the deleted base and the one after
  data.frame(AfterDeletion = aln$ref[single_del_locs + 1], 
             Deleted = aln$ref[single_del_locs])
}
# Apply our function to investigate deletions to all alignments
deletions = lapply(allAlns, findSingleDeletions)
# Combine the results
dels = do.call(rbind, deletions)
# And examine the table
knitr::kable(table(dels$AfterDeletion, dels$Deleted))

# Tables can also be visualized as heatmaps
dels$Count = 1
agg = aggregate(Count ~ AfterDeletion + Deleted, dels, length )
ggplot(agg, aes(x=AfterDeletion, y=Deleted, fill=Count)) + geom_tile() +
  theme_bw(base_size=10)  + 
  labs(title="Deletion by Context",
       x="Base After Deletion", y="Deleted Base")
```  
  
  
## More Advanced Alignment Metrics

pbbamr also enables you to see valuable metrics like SNR, pkMid, the start frame
and the pulse width if they are available in the alignment file (as they are in
internal mode BAMS).

```{r, echo=TRUE, results='asis'}
ifastaname = system.file("extdata", "All4Mer.V2.11.fna", package="pbbamr")
ibamname = system.file("extdata", "internalsample.bam", package="pbbamr")
# Load the index table.
iind = loadPBI(ibamname)
allAlns = loadAlnsFromIndex(iind, ifastaname)
# let's combine the individual alignments into one big data frame
alns = do.call(rbind, allAlns)
knitr::kable(head(alns[1:6,]))
ggplot(alns[alns$pkmid!=0,], aes(x=pkmid, fill=read)) + geom_density(alpha=.5) + theme_classic()
```
  
## Getting header information

You can obtain most of the information available in the header simply by calling
the loadHeader function.
```{r, echo=TRUE, results='asis'}
header = loadHeader(ibamname)
# Print out some of the items available in the list
header$version
knitr::kable(header$readgroups)
knitr::kable(header$sequences)
knitr::kable(header$programs)

```

